---
title: "Reproducible Research Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Loading and taking a look at the data summary.
```{r}
unzip("repdata_data_activity.zip")
if(!exists("activity")){
  activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
}
str(activity)
summary(activity)
```

##Processing or transforming the data into a format suitable for analysis.  Transform "steps" to numeric variable.  Transform "date" to POSIXct class using lubridate package.  
```{r}
activity$steps <- as.numeric(activity$steps)
library(lubridate)
activity$date <- ymd(activity$date)
str(activity)
```

## Question 1.  What is the average daily activity pattern?

### 1. Calculate the total number of steps taken per day (ignore the missing values)
```{r}
library(dplyr)
total_steps <- activity %>% group_by(date) %>% summarize(total=sum(steps, na.rm=TRUE)) %>% print
```

###  2. Make a histogram of the total number of steps taken each day
```{r}
hist(total_steps$total,breaks=20,main="Total steps per day",xlab="Steps per day", ylab="Number of Days")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean_steps <- mean(total_steps$total, na.rm=TRUE)
print(mean_steps)
median_steps <- median(total_steps$total, na.rm=TRUE)
print(median_steps)
```
The mean total steps is `r mean_steps`; the median total steps is `r median_steps`.

## Question 2.  What is the average daily activity pattern?

### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  
```{r}
average <- activity %>% group_by(interval) %>% summarise(average_steps=mean(steps, na.rm=TRUE))
plot(average$interval,average$average_steps, type="l", xlab="5 minute interval", ylab="average steps")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_interval <- average %>% filter(average_steps==max(average_steps))
print(max_interval$interval)
```
Interval `r max_interval$interval` contains the maximum number of steps.

## Question 3.  Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
missing <- sum(is.na(activity))
print(missing)
```
The total number of missing values in the dataset is `r missing`.

### 2. Devise a strategy for filling in all of the missing values in the dataset

The missing values are imputed using the average number of steps in that 5 minute interval.

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
steps_with_imputed <- numeric(nrow(activity))

for (i in 1:nrow(activity)) {
    if(is.na(activity[i,"steps"])==TRUE)
    { steps_with_imputed[i] <- average %>% filter(interval == activity[i,"interval"]) %>% select(average_steps)
    }
    else{
      steps_with_imputed[i] <- activity[i,"steps"]
    }
}
activity_with_imputed <- mutate(activity, steps=as.numeric(steps_with_imputed))
str(activity_with_imputed)
```

### 4. Make a histogram of the total number of steps taken each day 
```{r}
library(ggplot2)
total_steps_imputed <- tapply(activity_with_imputed$steps,activity_with_imputed$date,sum)

qplot(total_steps_imputed,main="Total steps per day",xlab="Steps per day", ylab="Number of Days",binwidth=500)
```


### 5. Calculate and report the mean and median total number of steps taken per day
```{r}
total_imputed <- activity_with_imputed %>% group_by(interval) %>% summarise(total=sum(steps))

mean_imputed <- mean(total_imputed$total)
print(mean_imputed)

median_imputed <- median(total_imputed$total)
print(median_imputed)
```
The mean total steps is `r mean_imputed`, which is less than the previously estimated mean total steps, 9354.23; the median total steps is `r median_imputed`, which is less than the previously estimated median total steps, 10395.  Imputing missing data using the average steps per interval resulted to a smaller estimate of the total daily number of steps.

## Question 4.  Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels indicating whether a given date is a weekday or weekend day
```{r}
activity_with_imputed_week <- activity_with_imputed %>% mutate(week=weekdays(activity_with_imputed$date)) %>% mutate(weekDay = ifelse(week==c("Saturday", "Sunday"),"weekend","weekday"))
```

### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
weekend <- tapply(activity_with_imputed_week[activity_with_imputed_week$weekDay=="weekend",]$steps, activity_with_imputed_week[activity_with_imputed_week$weekDay=="weekend",]$interval,mean)
weekday <- tapply(activity_with_imputed_week[activity_with_imputed_week$weekDay=="weekday",]$steps, activity_with_imputed_week[activity_with_imputed_week$weekDay=="weekday",]$interval,mean)

par(mfrow=c(1,2))
plot(as.numeric(names(weekend)),weekend,type="l", main="Average steps, weekends", xlab="Interval", ylab="Steps")
plot(as.numeric(names(weekday)),weekday,type="l", main="Average steps, weekday", xlab="Interval", ylab="Steps")
```

